% !TeX root = ../../thesis.tex
\chapter*{Preface}
\label{ch:preface}

After years of researching, laboring, ruminating, and questioning my life choices, the time has come to put everything in writing.
You hold in your hands the fortunate result of a series of questionable decisions, least of which wondering whether I should do a PhD.
This dissertation was made possible only through procrastination, dead-ends, frustration, creative bouts, self-doubt, and the occasional spark of clarity.

% What started as mild curiosity though turned into full-fledged fascination, and thus I present you the fruits of my frivolous endeavors.
If such milestones are meant to bring a sense of achievement or even entitlement, I like to think I am walking away from this a little wiser and a little humbler.
Being jubilant it is because of the gratitude I feel, so allow me this opportunity to express it.
Without my people, I would have lost my way long ago and this would have been nothing more than a collection of half-baked ideas and incomplete projects.
This dissertation is dedicated to all of you.

First and foremost, to Davy and Wouter, thank you for taking a gamble on me when you had no reason to. Your trust, acumen, advice, and support helped me channel what potential I had and for that, I am endlessly thankful.

To Stelios, for his enthusiasm, insightfulness, and determination are precious to this world. Also for being the single most important catalyst in my decision to pursue a PhD in Belgium. I will forever cherish our times of irreverence and camaraderie.

To Vera, for her trust and guidance that made me believe. You're not just an exemplary academic but a wonderful friend, and working with you has been a privilege. Your integrity and conscientiousness are an inspiration.

To all my dear friends and colleagues, that without you I would have amounted to nothing: Fotis, Paschalis, Christos, Konstantinos, Amr, Nikiforos, Sotiris, Christos, Jo, Job, Kshitij, Filippo, Giuseppe, Tim, Abel, Ali, Sander, Dimos, and all other incredible people that I could not possibly name all at once.
Your friendship have shaped me in ways you may never realize.

To Lieven Desmet, Hendrik Blockeel, Lorenzo Cavallaro and Maura Pintor for being inspiring mentors and accepting my invitation to be members of my committee.

To the DistriNet crew -- Annick Vandijck, Katrien Janssens, An Makowski, and everyone else: you have gone out of your way to support me, and I will never forget the great times we shared in and out of the department.

To Agnė, my partner in crime, for being that extraordinary nugget of brilliance, inquisitiveness, and compassion. Your appreciation and affection nurture me more than you could imagine. I cannot wait for our future adventures and to tackle life's challenges, starting from what is (its) meaning.

And most importantly to my parents, my sisters, and my grandparents.
Without your unwavering support, love, and commitment, I would have not made it this far. You are my greatest source of inspiration, and if I could choose a family, I would choose you every time. I am the luckiest person to have you.

For all the times I faltered, it was because of you all that I got back up and kept moving forward by forging new paths -- grateful for, and always striving to be worthy of the faith you placed in me.
So without further contrafibularities, it is time to draw the curtains on this one; now for something completely different!

\begin{flushright}
    \textit{-- Ilias Tsingenopoulos} \\
    \textit{October 2024, Leuven}
\end{flushright}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Keep the following \cleardoublepage at the end of this file, 
% otherwise \includeonly includes empty pages.
\cleardoublepage

% vim: tw=70 nocindent expandtab foldmethod=marker foldmarker={{{}{,}{}}}
